Fixing Vulnerabilities Potentially Hinders Maintainability

### 2021-02-08 paper/emse20.tex
pass on RQ3

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security patches that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.
% \end{itemize}
%
\subsection{Security Patches vs. Regular Changes}
%
Previous studies attempted to measure the impact of regular changes 
on open-source software maintainability~\cite{HEGEDUS2018313}. 
However, there is no previous work focused on comparing the impact 
of security patches with regular changes on maintainability, only
with bug-fixes~\cite{10.1145/3133956.3134072}.
We analyze the maintainability of regular changes---changes not 
related to security patches---and, use them as a baseline.
The baseline dataset is generated from the security commits dataset, i.e.,
for each security commit in the dataset, we collect a random regular
change from the same project. We created two different baselines: 
\textit{random-baseline},
considering random changes and all their characteristics; and,
\textit{size-baseline}, considering also random changes
but with an approximate size as security patches---we argue 
that comparing changes with considerable different sizes is unfair.

@@

As for the security patches, for each regular change, we  
need the commit performing the regular change (\emph{sha-reg}) and version of 
the software before the change (\emph{sha-reg-p}). For each security patch, a 
random commit from the same project is selected, \emph{sha-reg}.
The parent of \emph{sha-reg} or the commit performed before \emph{sha-reg} is \emph{sha-reg-p}.

@@

For the size-baseline, we also need to obtain the regular change (\emph{sha-reg})
and the version of the software before the change (\emph{sha-reg-p}). 
First, the \emph{diff}
between the security patch and its parent is calculated.
Then, a random commit/regular change from 
the same project is selected (\emph{sha-reg}). The 
\emph{diff} between \emph{sha-reg} and its parent (\emph{sha-reg-p})
is calculated and compared to the \emph{diff} from the security patch.
Due to the complexity of some patches, it was not possible 
to find patches with the exact number of added and deleted lines. 
Thus, we looked for an approximation.
% The algorithm integrates the following steps:

% \begin{enumerate}
% \item Calculates the \emph{diff} between the security patch commit 
% (\emph{sha}) and its parent (\emph{sha-p}).
% \item Selects a random commit from the project (\emph{sha-reg}).
% \item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) 
% and its parent (\emph{sha-reg-p}).
% \item Checks if the size from the regular change has the 
% same/approximated size to the security patch.
% \end{enumerate}

The pair of the regular change (\emph{sha-reg}) and its parent (\emph{sha-reg-p}) is accepted if the 
\emph{diff} size fits in the range size. This range widens every 
$10$ attempts to search for a change with the same size. We 
originate the regular changes from the security commits to ensure 
that differences in maintainability are not a consequence of 
characteristics of different projects.

@@

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of 
commits (e.g., baseline and security commits), we use the Paired 
Wilcoxon signed-rank test with the significance level $\alpha = 
0.05$~\cite{10.2307/3001968}. In other words, we test the null 
hypothesis that the maintainability difference between pairs of 
versions $v_{s-1}$, $v_s$ (i.e., before and after a security-commit) 
come from the same distribution. Nevertheless, this test 
has a limitation: it does not consider the groups of commits with a 
zero-difference maintainability. In $1959$, Pratt provided an 
improvement to the test to solve this issue making the test more 
robust. Thus, we use a version of the Wilcoxon test that 
incorporates the cases where maintainability is different from 
zero~\cite{10.2307/2282543}. The Wilcoxon test requires a 
distribution size of at least $20$ instances. To understand the 
effect-size, as advocated by the Common-language effect 
sizes, we compute the mean difference, the median of 
the difference, and the percentage of cases that reduce 
maintainability~\cite{graw:1992}.
%

@@

\textit{\textbf{RQ3: What is the impact of security changes versus regular 
changes on the maintainability of open-source software?}}
The impact of security and regular changes on software maintainability 
is presented in Figure~\ref{fig:secvsreg}. In this section, 
we present a comparison of security patches with 
two different baselines of regular changes: 
\emph{size-baseline}, a dataset of  
random regular changes with the same size as security 
patches---we argue that comparing changes with considerable different 
sizes may be unfair; and, \emph{random-baseline},
a dataset of random regular changes. Our hypothesis in this research question is
that \emph{Security patches hinder more software maintainability and 
need more attention than regular changes}.

We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.9\%$ ($406$) of patches suffered a 
negative impact, $38.7\%$ ($375$) of patches remained the same and $19.4\%$ 
($188$) of patches increased software maintainability. For regular changes,
when considering the size of the changes (\emph{size-basline}),
we observe that the maintainability decreases in $27.0\%$ ($262$) 
and increases in $30.5\%$ ($295$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.5\%$ ($412$) of the cases and perform regular
changes has a more positive impact than negative on  
maintainability. Fully regular changes (\emph{random-baseline}) 
are less prone to hinder software maintainability than 
security changes. About $34.4\%$ ($333$) of the regular changes
have a negative impact on software maintainability compared with 
the $41.9\%$ of security patches.

@@

Overall, the results for both baselines, show that regular 
changes are less prone to hinder the software 
maintainability of open-source software.   
We manually inspected a considerable amount of cases from
the distribution of regular changes that considers size (\emph{size-baseline}), 
where the impact on maintainability 
increases, and, found that identifying regular changes with the same 
size as the security-related commit is limiting the type of regular 
commits being randomly chosen: input 
patches, variables or functions, type conversion, i.e., changes
with no impact on the software metrics BCh evaluates.
We presume that this phenomenon lead
to the significant number of cases where there is no impact
on the software maintainability. On the other hand, 
identifying regular changes without any restrictions (\emph{random-baseline})
shows that regular changes have less negative impact 
on software maintainability compared with security patches
with statistical significance ($p = 5.34$x$10^{-8}$) and that 
special attention should be given to security patches.

\textbf{Summary:} Security-related commits are observed to harm software 
maintainability. While regular changes with no characteristics restrictions 
show that regular changes are less prone to harm software maintainability.
Thus, we urge the importance of adopting maintainability practices while 
applying security patches.

>>>>

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security patches that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.
% \end{itemize}
%
\subsection{Security Patches vs. Regular Changes}
%
Previous studies attempted to measure the impact of regular changes 
on open-source software maintainability~\cite{HEGEDUS2018313}. 
However, there is no previous work focused on comparing the impact 
of security patches with regular changes on maintainability, only
with bug-fixes~\cite{10.1145/3133956.3134072}.
We analyze the maintainability of regular changes---changes not 
related to security patches---and, use them as a baseline.
The baseline dataset is generated from the security commits dataset, i.e.,
for each security commit in the dataset, we collect a random regular
change from the same project. We created two different baselines: 
\textit{random-baseline},
considering random changes and all their characteristics; and,
\textit{size-baseline}, considering also random changes
but with an approximate size as security patches---we argue 
that comparing changes with considerable different sizes may be unfair.

@@

As for the security patches, for each regular change, we  
need the commit performing the regular change (\emph{sha-reg}) and version of 
the software before the change (\emph{sha-reg-p}). For each security patch, a 
random commit from the same project is selected, \emph{sha-reg}. The parent commit
of \emph{sha-reg} is the \emph{sha-reg-p}.

@@

For the size-baseline, we also need to obtain the regular change (\emph{sha-reg})
and the version of the software before the change (\emph{sha-reg-p}). 
First, our tool calculates the \emph{diff}
between the security patch and its parent.
Second, a random commit/regular change from 
the same project is selected, \emph{sha-reg}. The 
\emph{diff} between \emph{sha-reg} and its parent (\emph{sha-reg-p})
is calculated. Then, the regular change \emph{diff} is compared 
to the security patch \emph{diff}.
Due to the complexity of some patches, it was not possible 
to find patches with the exact same number of added and deleted lines. 
Thus, we looked for an approximation.

The pair of the regular change (\emph{sha-reg}) and its parent (\emph{sha-reg-p}) is accepted if the 
\emph{diff} size fits in the range size. This range widens every 
$10$ attempts to search for a change with an approximate size. We 
originate the regular changes from the security commits to ensure 
that differences in maintainability are not a consequence of 
characteristics of different projects.

@@

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of 
commits (e.g., baseline and security commits), we use the Paired 
Wilcoxon signed-rank test with the significance level $\alpha = 
0.05$~\cite{10.2307/3001968}. In other words, we test the null 
hypothesis that the maintainability difference between pairs of 
versions $v_{s-1}$, $v_s$ (i.e., before and after a security-commit) 
come from the same distribution. Nevertheless, this test 
has a limitation: it does not consider the groups of commits with a 
zero-difference maintainability. In $1959$, Pratt provided an 
improvement to the test to solve this issue making the test more 
robust. Thus, we use a version of the Wilcoxon test that 
incorporates the cases where maintainability is equal to 
zero~\cite{10.2307/2282543}. The Wilcoxon test requires a 
distribution size of at least $20$ instances. To understand the 
effect-size, as advocated by the Common-language effect 
sizes, we compute the mean difference, the median of 
the difference, and the percentage of cases that reduce 
maintainability~\cite{graw:1992}.
%

@@

\textit{\textbf{RQ3: What is the impact of security patches versus regular 
changes on the maintainability of open-source software?}}
The impact of security and regular changes on software maintainability 
is presented in Figure~\ref{fig:secvsreg}. In this section, 
we present a comparison of security patches with 
two different baselines of regular changes: 
\emph{size-baseline}, a dataset of  
random regular changes with the same size as security 
patches---we argue that comparing changes with considerable different 
sizes may be unfair; and, \emph{random-baseline},
a dataset of random regular changes.

Our hypothesis is
that \emph{security patches hinder more software maintainability 
than regular changes}. 
We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.9\%$ ($406$) of patches suffered a 
negative impact, $38.7\%$ ($375$) of patches remained the same and $19.4\%$ 
($188$) of patches increased software maintainability. For regular changes,
when considering the size of the changes (\emph{size-basline}),
we observe that the maintainability decreases in $27.0\%$ ($262$) 
and increases in $30.5\%$ ($295$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.5\%$ ($412$) of the cases and perform regular
changes has a more positive impact than negative on  
maintainability. However, no statistical significance was 
obtained. Regular changes (\emph{random-baseline}), with no size 
restrictions, are less prone to hinder software maintainability than 
security changes. About $34.4\%$ ($333$) of the regular changes 
have negative impact on software maintainability---less than 
in the security patches. For the \emph{random-baseline}, 
statistical significance was retrieved ($p = 5.34$x$10^{-8}$).

@@

Overall, the results for both baselines, show that regular 
changes are less prone to hinder the software 
maintainability of open-source software.   
We manually inspected a considerable amount of cases from
the distribution of regular changes that considers size (\emph{size-baseline}), 
where the impact on maintainability 
increases, and, found that identifying regular changes with the same 
size as the security-related commit is limiting the type of regular 
commits being randomly chosen: input 
patches, variables or functions, type conversion, i.e., changes
with no impact on the software metrics analyzed by BCH.
We presume that this phenomenon lead
to the significant number of cases where there is no impact
on the software maintainability. On the other hand, 
identifying regular changes without any restrictions (\emph{random-baseline})
shows that regular changes have less negative impact 
on software maintainability compared with security patches
and that special attention should be given to security patches.

\textbf{Summary:} Security-related commits are observed to harm software 
maintainability, while regular changes are less prone to harm software maintainability.
Thus, we urge the importance of adopting maintainability practices while 
applying security patches.
